Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes

### 2024-12-15 paper.tex
other files

---
title: "Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes"
author: "Rishika Randev, Jenny Wu, Uzoma Uwazurike Jr., Shiyue Zhou"
instructor: "Andrea Lane"
institution: "Duke University, IDS 702 Data Modeling and Presentation"
format: pdf
header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
fontsize: 10pt
editor: visual
execute:
  echo: false
  warning: false
---

```{r, echo= FALSE, warning = FALSE, message=FALSE}
library("dplyr") 
library(mice) 
library(sjlabelled) 
library(tidyverse) 
library("modelsummary") 
library(car)
library(GGally)
library(ggplot2)
library(patchwork)
library(caret)
library(pROC)
library(pander)
library(kableExtra)
```

```{r,echo= FALSE, warning = FALSE, message=FALSE}
combined <- read.csv('https://github.com/wujenny214/F24_STATS_FINAL/raw/refs/heads/main/combined.csv')

```

```{r, echo= FALSE, warning = FALSE}
# Making the state_fac variable a factor variable for future use
combined <- combined |> mutate(State_fac = factor(combined$State), Region_fac = factor(combined$Region, levels=c('South', 'West', 'Northeast', 'Midwest')))

```

```{r, echo= FALSE, warning = FALSE}
## Imputation Code 
# In this block, we are looking to use predictive mean matching to combing regression methods and hot deck imputing to predicted value closest to the predicted value of the observation with missing data; we also needed to convert school funding adequacy imputed values to school funding categorical variable before using in models

sub <- combined |> select(c("Life.Expectancy","Residential.Segregation.Index","X..Limited.Access.to.Healthy.Foods","X..Uninsured.Adults","Median.Household.Income", "School.Funding.Adequacy", "Median.Prop.Tax", "Region_fac"))
sub <- unlabel(sub) #unlabel the data (labels cause problem for the mice function)

# Add School.Funding.Cat to the orginal dataset. Since we do not want to imputate on the categorical outcome variable,  this creation of a variable exists to create the proper dataframe within the imputated data 
sub$School.Funding.Cat <- ifelse(sub$School.Funding.Adequacy > 0, 1, 0)

# Perform the imputation
sub.imp <- mice(sub, m = 1, method = "pmm", print = FALSE, seed=3)

```

```{r, echo= FALSE, warning = FALSE}
# Since we should not imputate on the outcome, but need to imputate the School.Funding.Adequacy variable, we are feeding the School.Funding.Adequacy variable into a ifelse conditional that will replaced the imputed outcome variable.

# Define the categorization function
categorize_school_funding <- function(x) {
  ifelse(x > 0, 1, 0)  # Binary classification
}

# Loop over the imputations
for (i in 1:1) {
  # Extract the imputed values for the i-th dataset
  imputed_values <- sub.imp$imp$School.Funding.Adequacy[, i]

  # Categorize the imputed values
  categorized_values <- categorize_school_funding(imputed_values)
  complete_cat <- rep(NA, nrow(sub))  # Initialize with NAs

  # Get the indices where imputation occurred
  imp_indices <- which(!is.na(categorized_values))  # Only for imputed values

  # Assign the categorized values to the rows that were imputed
  complete_cat[imp_indices] <- categorized_values[imp_indices]

  # Save into the `imp` list of the `sub.imp` object
  sub.imp$imp$School.Funding.Cat[[i]] <- complete_cat[imp_indices]
}

```

```{r, echo= FALSE, warning = FALSE, include = FALSE }
# To have a full dataset, we are taking the first imputed dataset R provided and using that as our full dataset sample. 
completed_data <- complete(sub.imp, 1)

# Check the structure
str(completed_data)

# Ensure that there are no missing data 
colSums(is.na(completed_data))
```

## Abstract

@@

## **Introduction**

Life expectancy and school funding are important indicators of public well-being because they capture fundamental aspects of health, education, and social equity that are essential for the development of individuals and communities. Life expectancy is an important health indicator that reflects not only the overall health of a society, but also its stability, economic resilience, and levels of inequality. While access to quality healthcare and nutritious foods are drivers of well-being, where you live and the nature of the community you live in likely also have an influence on health outcomes such as life expectancy. According to "The Root Causes of Health Inequity - Communities in Action" report from the National Institutes of Health, we can infer that differences in life expectancy between regions and socioeconomic groups often reveal systemic inequalities, with marginalized communities experiencing shorter life spans due to limited access to health care, healthy living environments, and economic opportunities.^1^ When it comes to educational indicators, adequately funded schools ensure that students have equal access to learning opportunities, serving as a crucial investment in social mobility. The Stanford Center for Opportunity Policy in Education highlights that "the unequal allocation and inadequate levels of resources in schools and communities \[are\] at the heart of many gaps in student opportunity."^2^ This emphasizes the importance of equitable and adequate school funding in closing opportunity gaps and empowering students from disadvantaged backgrounds to achieve academic and economic success. By better understanding the correlates of school funding adequacy, we can take a major step towards implementing policy that targets the areas and communities most affected by this issue.

For this purpose, we focus our analysis on two key research questions. First, we investigate how socioeconomic factors impact average life expectancy. **We seek to answer the question, "What is the impact of residential segregation, access to healthy foods, and health insurance coverage on a county's average life expectancy?".** This analysis includes a comparison across different regions to identify variations in the relationship between residential segregation and life expectancy. Second, **we seek to answer the question, "Do residential segregation and region impact the likelihood of a school being adequately funded?"**. This question allows us to assess how the racial makeup and geographic location—given that each region of the U.S. is characterized by differing political and cultural norms—affect a county's educational investment. The observational unit for this analysis is the U.S. county.

To obtain data on the aforementioned variables, we used three primary sources.

The first source was the County Health Rankings & Roadmaps program,^3^ associated with the University of Wisconsin Population Health Institute, which provided us with 2024 datasets for all U.S. states. This program consolidates the latest county-level measurements of various population health, economic, demographic, and social factors from the American Community Survey, the USDA Food Environment Atlas, Small Area Income and Poverty Estimates, and other reliable government sources into publicly available datasets every year. Our specific variables of interest were originally collected as part of:

-   National Center for Health Statistics - Natality & Mortality Files, 2019-2021 (average life expectancy)

-   American Community Survey 5-year estimates, 2018-2022 (residential segregation index & median household income)

-   USDA Food Environment Atlas, 2019 (percentage of population with limited access to healthy foods)

-   Small Area Health Insurance Estimates, 2021 (percentage of adults uninsured)

-   School Finance Indicators Database, 2021 (school funding adequacy)

The second source was IPUMS, which provided us with median poverty tax data for every county from 2018-2022.^4^

The third source was the U.S. Census Bureau’s Regions and Divisions of the United States, which maps every U.S. to one of four geographic regions, and also a division within the regions. We specifically included regions as a variable in this analysis. This data was loaded into R using a [publicly available GitHub repository](https://github.com/cphalpert/census-regions) titled *Census Regions*.^5^

## **Methods**

### Variable Selection

Variables of interest were selected a priori based on their perceived relevance to the two outcomes we wanted to better understand: average life expectancy and school funding adequacy. While nutrition and access to quality healthcare are inherently linked to general wellbeing, it is worthwhile to assess how influential the environmental and financial aspects of these factors are on our tangible outcome. Therefore, we selected two key predictors for our life expectancy model: the percentage of the population with limited access to healthy foods and the percentage of uninsured adults in a county. The former is defined by the US Department of Agriculture as "the percentage of the \[county\] population that is low-income and does not live close to a grocery store", where "close" is defined differently for rural and nonrural areas.^6^ The latter variable represents the percentage of adults in a county younger than 65 who lack health insurance. We also were interested in seeing how residential segregation would impact life expectancy among these predictors, given existing studies on the negative health effects of segregation, and whether this relationship would exhibit any variations across different regions. This is why we chose to include residential segregation index values as a predictor. The residential segregation index, used by the County Health Rankings & Roadmaps program, measures how evenly two groups, such as Black and white residents, are distributed across geographic areas, with values ranging from 0 (complete integration) to 100 (complete segregation).^7^

Past studies have established a correlation between income and life expectancy.^8^ Because we wanted to control for its effects and examine how insurance, healthy food access, segregation, and region influence life expectancy even within a given income bracket, we selected income as our final predictor for the first model.

School funding adequacy is defined by the County Health Rankings program as "the average dollar gap between actual per-pupil spending and the amount needed for students to achieve national average test scores, considering districts' varying equity-based needs."^9^ To explore how racial diversity interface with variations in school funding, we selected residential segregation to be a part of this regression. In addition, given that state policy has a large influence on school administration and the determination of school needs, and that states in the same region tend to have similar political leanings and cultural beliefs, we wanted to analyze how much of a relationship exists between region and actual school funding levels. Because a major source of public school funding is tax revenue, median property tax and median household income were also included in the school funding model as potential confounders. This allowed us to assess whether the effects of segregation and region on funding existed outside of the effects of known economic factors.

### Exploratory Data Analysis & Data Preprocessing

2024 datasets for all U.S. states were combined and merged with median property tax and region mapping data, resulting in a final dataset with 3144 county observations and 9 different variables (state, average life expectancy, residential segregation index, % limited access to healthy foods, % uninsured adults, school funding adequacy, median household income, median property tax, and region). Because school funding adequacy was given as a numerical value, and we wanted to focus on better understanding the dichotomy between school funding adequacy and inadequacy, this variable was converted into a binary categorical variable (where the value was set to 1 if the numerical value was positive, indicating the counties’ schools were adequately funded, and set to 0 if the numerical value was negative, indicating underfunding).

Exploratory data analysis for the two outcome variables was conducted by 1) calculating descriptive statistics, 2) creating scatterplots of each individual predictor against the outcome (for life expectancy), and 3) creating pairwise plots between variables (for school funding). During the initial stages of analysis, we found that ⅓ of our observations were missing residential segregation index data, and that certain states were missing nearly all of their segregation values. Because of this, we decided to drop all states missing over 50% of their values for any variable. This led to our analysis being limited to 33 states, and excluding the following: Alaska, Colorado, Idaho, Iowa, Kansas, Minnesota, Montana, Nebraska, North Dakota, South Dakota, Utah, Wyoming, and Vermont. We were left with 2379 total observations, and the missing percentage for residential segregation dropped to 22%.

@@

### Model Fitting & Assessment

For our first research question, a multiple linear regression model was fit, regressing average life expectancy on residential segregation index, percentage with limited access to healthy foods, percentage uninsured adults, region, and median household income. Linear regression assumptions were assessed using diagnostic plots, especially residual vs. fitted and quantile-quantile plots. The adjusted R-squared value was used to evaluate the fit of the model, and the effect of region as an interaction term with residential segregation was evaluated using nested F tests. We also calculated VIF to check for multicollinearity and Cook’s distance to identify influential points.

@@

## Results

### *Data Overview*

Table 1 displays summary statistics for the predictors in the second model, stratified by school funding adequacy. Table A1 in the Appendix provides summary statistics for all variables, and shows that both the mean and median for life expectancy are approximately 75 years. School funding adequacy counts show that around 66% of the counties in our analysis had underfunded schools. Additionally, the number of counties per region indicate an imbalance between categories, with the South region being over represented (holding 60% of all counties), and the East and Northeast holding around 9% of the counties each.

```{r, echo=FALSE, include=FALSE}
completed_data |> group_by(School.Funding.Cat) |> summarise(mean(Median.Household.Income), sd(Median.Household.Income), mean(Median.Prop.Tax), sd(Median.Prop.Tax), mean(Residential.Segregation.Index), sd(Residential.Segregation.Index))
```

```{r, echo=FALSE, include=FALSE}
table(completed_data$School.Funding.Cat, completed_data$Region_fac)
```

@@

*Table 1:* *School Funding Adequacy Summary Statistics*

| Variable                                 | **Inadequate School Funding** | **Adequate School Funding** |
|------------------------------|----------------------|--------------------|
| Residential Segregation Index: Mean (SD) | 46 (17)                       | 56 (14)                     |
| Region-West: Count (%)                   | 114 (7.3%)                    | 89 (10.8%)                  |
| Region-Northeast: Count (%)              | 3 (0.2%)                      | 200 (24.3%)                 |
| Region-Midwest: Count (%)                | 267 (17.1%)                   | 285 (34.7%)                 |
| Region-South: Count (%)                  | 1173 (75.3%)                  | 248 (30.2%)                 |
| Median Property Tax, USD: Mean (SD)      | 1119 (763)                    | 2282 (1648)                 |
| Median Household Income, USD: Mean (SD)  | 56513 (13200)                 | 72218 (18255)               |

```{r, echo=FALSE, include=FALSE}
# Plot for Q2 EDA
ggpairs(
  completed_data,
  columns = c("Residential.Segregation.Index", "Median.Household.Income", "Median.Prop.Tax"),
  aes(color = factor(School.Funding.Cat,
                     labels = c("Inadequate", "Adequate")))
) +
  labs(
    title = "Pairwise Plots of Predictors Colored by School Funding Adequacy",
    color = "School Funding Adequacy"
  ) +
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    strip.text.x = element_text(size = 5),
    strip.text.y = element_text(size = 3)
  )

```

### Question 1

#### Model Results

@@

Surprisingly, holding all other variables constant, for every 1 unit increase in percentage of uninsured adults in a county, a county's average life expectancy increases by 0.032 years. While statistically significant, many studies covering the relationship between an individual’s health and having insurance coverage have proven its positive correlation, suggesting our results may benefit from a closer look. In assessing our dataset, as seen in Appendix 1A, we find that our dataset contains more observations for counties with low uninsured individuals, or equivalently, counties where most people are covered. As the timeframe of our dataset is 2018-2022, it is likely that the aforementioned observations require greater scrutiny of the significance of the impact of this variable on average life expectancy of a county.

@@

*Figure 1: Segregation vs. Life Expectancy Across Different U.S. Regions*

```{r, echo=FALSE, fig.pos='H'}
library(ggplot2)
library(GGally)
library(patchwork)

ggplot(completed_data, aes(x = Residential.Segregation.Index, y = Life.Expectancy, color=Region_fac)) +
  geom_point(color='black') + geom_smooth(method="lm",se=F) + 
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Residential Segregation Index", y = "Average Life Expectancy (Years)", color='U.S. Region', title='Relationship Between Residential Segregation and Life Expectancy')+ 
  theme(plot.margin = margin(t = 20, r = 30, b = 20, l = 10),
axis.title.x = element_text(size = 8),
axis.title.y = element_text(size = 8),
title=element_text(size=9))
```

*Table 2: MLR Results For Average Life Expectancy Outcome*

```{r, echo= FALSE, warning = FALSE, fig.pos='H'}
# Running a regression with the imputated data (completed_data) to understand the impact of missing data

mlr.imp <- lm(Life.Expectancy ~ Residential.Segregation.Index * Region_fac +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults + Median.Household.Income, data=completed_data)

renamed <- c(
    "Residential.Segregation.Index" = "Residential Segregation Index (RSI)",
  "X..Limited.Access.to.Healthy.Foods" = "Percentage With Access to Healthy Foods",
  "X..Uninsured.Adults"= "Percentage of Uninsured Adults",
  "Region_facWest"="Region-West", 
  "Region_facNortheast"="Region-Northeast",
  "Region_facMidwest"="Region-Midwest" ,
  "Residential.Segregation.Index:Region_facWest" = "RSI*Region-West",
  "Residential.Segregation.Index:Region_facNortheast"= "RSI*Region-Northeast",
  "Residential.Segregation.Index:Region_facMidwest"="RSI*Region-Midwest",
  "Median Household Income" = "Median.Household.Income")

modelsummary(mlr.imp, fmt = fmt_significant(2),
             coef_rename = renamed, 
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA,
             )
```

```{r echo= FALSE, warning = FALSE, include=FALSE}
# Running a regression without the interaction term between residential segregation and region

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{rotating}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\KOMAoption{captions}{tableheading}
\usepackage{float}
\floatplacement{table}{H}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes},
  pdfauthor={Rishika Randev, Jenny Wu, Uzoma Uwazurike Jr., Shiyue Zhou},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Understanding the Impact of Socioeconomic and Regional Factors on
Health and Educational Outcomes}
\author{Rishika Randev, Jenny Wu, Uzoma Uwazurike Jr., Shiyue Zhou}
\date{}

\begin{document}
\maketitle

\subsection{Abstract}\label{abstract}

@@

\subsection{\texorpdfstring{\textbf{Introduction}}{Introduction}}\label{introduction}

Life expectancy and school funding are important indicators of public
well-being because they capture fundamental aspects of health,
education, and social equity that are essential for the development of
individuals and communities. Life expectancy is an important health
indicator that reflects not only the overall health of a society, but
also its stability, economic resilience, and levels of inequality. While
access to quality healthcare and nutritious foods are drivers of
well-being, where you live and the nature of the community you live in
likely also have an influence on health outcomes such as life
expectancy. According to ``The Root Causes of Health Inequity -
Communities in Action'' report from the National Institutes of Health,
we can infer that differences in life expectancy between regions and
socioeconomic groups often reveal systemic inequalities, with
marginalized communities experiencing shorter life spans due to limited
access to health care, healthy living environments, and economic
opportunities.\textsuperscript{1}~When it comes to educational
indicators, adequately funded schools ensure that students have equal
access to learning opportunities, serving as a crucial investment in
social mobility. The Stanford Center for Opportunity Policy in Education
highlights that ``the unequal allocation and inadequate levels of
resources in schools and communities {[}are{]} at the heart of many gaps
in student opportunity.''\textsuperscript{2} This emphasizes the
importance of equitable and adequate school funding in closing
opportunity gaps and empowering students from disadvantaged backgrounds
to achieve academic and economic success. By better understanding the
correlates of school funding adequacy, we can take a major step towards
implementing policy that targets the areas and communities most affected
by this issue.

For this purpose, we focus our analysis on two key research questions.
First, we investigate how socioeconomic factors impact average life
expectancy. \textbf{We seek to answer the question, ``What is the impact
of residential segregation, access to healthy foods, and health
insurance coverage on a county's average life expectancy?''.} This
analysis includes a comparison across different regions to identify
variations in the relationship between residential segregation and life
expectancy. Second, \textbf{we seek to answer the question, ``Do
residential segregation and region impact the likelihood of a school
being adequately funded?''}. This question allows us to assess how the
racial makeup and geographic location---given that each region of the
U.S. is characterized by differing political and cultural norms---affect
a county's educational investment. The observational unit for this
analysis is the U.S. county.

To obtain data on the aforementioned variables, we used three primary
sources.~

The first source was the County Health Rankings \& Roadmaps
program,\textsuperscript{3} associated with the University of Wisconsin
Population Health Institute, which provided us with 2024 datasets for
all U.S. states. This program consolidates the latest county-level
measurements of various population health, economic, demographic, and
social factors from the American Community Survey, the USDA Food
Environment Atlas, Small Area Income and Poverty Estimates, and other
reliable government sources into publicly available datasets every year.
Our specific variables of interest were originally collected as part of:

\begin{itemize}
\item
  National Center for Health Statistics - Natality \& Mortality Files,
  2019-2021 (average life expectancy)
\item
  American Community Survey 5-year estimates, 2018-2022 (residential
  segregation index \& median household income)
\item
  USDA Food Environment Atlas, 2019 (percentage of population with
  limited access to healthy foods)
\item
  Small Area Health Insurance Estimates, 2021 (percentage of adults
  uninsured)
\item
  School Finance Indicators Database, 2021 (school funding adequacy)
\end{itemize}

The second source was IPUMS, which provided us with median poverty tax
data for every county from 2018-2022.\textsuperscript{4}

The third source was the U.S. Census Bureau's Regions and Divisions of
the United States, which maps every U.S. to one of four geographic
regions, and also a division within the regions. We specifically
included regions as a variable in this analysis.~This data was loaded
into R using a
\href{https://github.com/cphalpert/census-regions}{publicly available
GitHub repository} titled \emph{Census Regions}.\textsuperscript{5}

\subsection{\texorpdfstring{\textbf{Methods}}{Methods}}\label{methods}

\subsubsection{Variable Selection}\label{variable-selection}

Variables of interest were selected a priori based on their perceived
relevance to the two outcomes we wanted to better understand: average
life expectancy and school funding adequacy. While nutrition and access
to quality healthcare are inherently linked to general wellbeing, it is
worthwhile to assess how influential the environmental and financial
aspects of these factors are on our tangible outcome. Therefore, we
selected two key predictors for our life expectancy model: the
percentage of the population with limited access to healthy foods and
the percentage of uninsured adults in a county. The former is defined by
the US Department of Agriculture as ``the percentage of the {[}county{]}
population that is low-income and does not live close to a grocery
store'', where ``close'' is defined differently for rural and nonrural
areas.\textsuperscript{6} The latter variable represents the percentage
of adults in a county younger than 65 who lack health insurance. We also
were interested in seeing how residential segregation would impact life
expectancy among these predictors, given existing studies on the
negative health effects of segregation, and whether this relationship
would exhibit any variations across different regions. This is why we
chose to include residential segregation index values as a predictor.
The residential segregation index, used by the County Health Rankings \&
Roadmaps program, measures how evenly two groups, such as Black and
white residents, are distributed across geographic areas, with values
ranging from 0 (complete integration) to 100 (complete
segregation).\textsuperscript{7}~

Past studies have established a correlation between income and life
expectancy.\textsuperscript{8} Because we wanted to control for its
effects and examine how insurance, healthy food access, segregation, and
region influence life expectancy even within a given income bracket, we
selected income as our final predictor for the first model.

School funding adequacy is defined by the County Health Rankings program
as ``the average dollar gap between actual per-pupil spending and the
amount needed for students to achieve national average test scores,
considering districts' varying equity-based needs.''\textsuperscript{9}
To explore how racial diversity interface with variations in school
funding, we selected residential segregation to be a part of this
regression. In addition, given that state policy has a large influence
on school administration and the determination of school needs, and that
states in the same region tend to have similar political leanings and
cultural beliefs, we wanted to analyze how much of a relationship exists
between region and actual school funding levels. Because a major source
of public school funding is tax revenue, median property tax and median
household income were also included in the school funding model as
potential confounders. This allowed us to assess whether the effects of
segregation and region on funding existed outside of the effects of
known economic factors.

\subsubsection{Exploratory Data Analysis \& Data
Preprocessing}\label{exploratory-data-analysis-data-preprocessing}

2024 datasets for all U.S. states were combined and merged with median
property tax and region mapping data, resulting in a final dataset with
3144 county observations and 9 different variables (state, average life
expectancy, residential segregation index, \% limited access to healthy
foods, \% uninsured adults, school funding adequacy, median household
income, median property tax, and region). Because school funding
adequacy was given as a numerical value, and we wanted to focus on
better understanding the dichotomy between school funding adequacy and
inadequacy, this variable was converted into a binary categorical
variable (where the value was set to 1 if the numerical value was
positive, indicating the counties' schools were adequately funded, and
set to 0 if the numerical value was negative, indicating underfunding).

Exploratory data analysis for the two outcome variables was conducted by
1) calculating descriptive statistics, 2) creating scatterplots of each
individual predictor against the outcome (for life expectancy), and 3)
creating pairwise plots between variables (for school funding). During
the initial stages of analysis, we found that ⅓ of our observations were
missing residential segregation index data, and that certain states were
missing nearly all of their segregation values. Because of this, we
decided to drop all states missing over 50\% of their values for any
variable. This led to our analysis being limited to 33 states, and
excluding the following: Alaska, Colorado, Idaho, Iowa, Kansas,
Minnesota, Montana, Nebraska, North Dakota, South Dakota, Utah, Wyoming,
and Vermont. We were left with 2379 total observations, and the missing
percentage for residential segregation dropped to 22\%.

@@

\subsubsection{Model Fitting \&
Assessment}\label{model-fitting-assessment}

For our first research question, a multiple linear regression model was
fit, regressing average life expectancy on residential segregation
index, percentage with limited access to healthy foods, percentage
uninsured adults, region, and median household income. Linear regression
assumptions were assessed using diagnostic plots, especially residual
vs.~fitted and quantile-quantile plots. The adjusted R-squared value was
used to evaluate the fit of the model, and the effect of region as an
interaction term with residential segregation was evaluated using nested
F tests. We also calculated VIF to check for multicollinearity and
Cook's distance to identify influential points.~

@@

\subsection{Results}\label{results}

\subsubsection{\texorpdfstring{\emph{Data
Overview}}{Data Overview}}\label{data-overview}

Table 1 displays summary statistics for the predictors in the second
model, stratified by school funding adequacy. Table A1 in the Appendix
provides summary statistics for all variables, and shows that both the
mean and median for life expectancy are approximately 75 years. School
funding adequacy counts show that around 66\% of the counties in our
analysis had underfunded schools. Additionally, the number of counties
per region indicate an imbalance between categories, with the South
region being over represented (holding 60\% of all counties), and the
East and Northeast holding around 9\% of the counties each.

@@

\emph{Table 1:} \emph{School Funding Adequacy Summary Statistics}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4167}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3056}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Inadequate School Funding}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Adequate School Funding}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Residential Segregation Index: Mean (SD) & 46 (17) & 56 (14) \\
Region-West: Count (\%) & 114 (7.3\%) & 89 (10.8\%) \\
Region-Northeast: Count (\%) & 3 (0.2\%) & 200 (24.3\%) \\
Region-Midwest: Count (\%) & 267 (17.1\%) & 285 (34.7\%) \\
Region-South: Count (\%) & 1173 (75.3\%) & 248 (30.2\%) \\
Median Property Tax, USD: Mean (SD) & 1119 (763) & 2282 (1648) \\
Median Household Income, USD: Mean (SD) & 56513 (13200) & 72218
(18255) \\
\end{longtable}

\subsubsection{Question 1}\label{question-1}

\paragraph{Model Results}\label{model-results}

@@

Surprisingly, holding all other variables constant, for every 1 unit
increase in percentage of uninsured adults in a county, a county's
average life expectancy increases by 0.032 years. While statistically
significant, many studies covering the relationship between an
individual's health and having insurance coverage have proven its
positive correlation, suggesting our results may benefit from a closer
look. In assessing our dataset, as seen in Appendix 1A, we find that our
dataset contains more observations for counties with low uninsured
individuals, or equivalently, counties where most people are covered. As
the timeframe of our dataset is 2018-2022, it is likely that the
aforementioned observations require greater scrutiny of the significance
of the impact of this variable on average life expectancy of a county.~

@@

\emph{Figure 1: Segregation vs.~Life Expectancy Across Different U.S.
Regions}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-10-1.pdf}

\emph{Table 2: MLR Results For Average Life Expectancy Outcome}

\begin{table}
\centering
\begin{tblr}[         %% tabularray outer open
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]Q[]Q[]Q[]},
cell{1}{2}={c=5,}{halign=c,},
column{1}={halign=l,},
column{2}={halign=c,},
column{3}={halign=c,},
column{4}={halign=c,},
column{5}={halign=c,},
column{6}={halign=c,},
row{1}={halign=c,},
}                     %% tabularray inner close
\toprule
& (1) &  &  &  &  \\ \cmidrule[lr]{2-6}
& Est. & S.E. & 2.5 \% & 97.5 \% & p \\ \midrule %% TinyTableHeader
(Intercept)                             & \num{65.88}    & \num{0.27}    & \num{65.34}    & \num{66.41}   & \num{<0.01} \\
Residential Segregation Index (RSI)     & \num{0.00715}  & \num{0.00332} & \num{0.00065}  & \num{0.01366} & \num{0.03}  \\
Region-West                             & \num{1.96}     & \num{0.69}    & \num{0.60}     & \num{3.32}    & \num{<0.01} \\
Region-Northeast                        & \num{2.92}     & \num{0.84}    & \num{1.27}     & \num{4.57}    & \num{<0.01} \\
Region-Midwest                          & \num{1.91}     & \num{0.44}    & \num{1.06}     & \num{2.77}    & \num{<0.01} \\
Percentage With Access to Healthy Foods & \num{0.0040}   & \num{0.0062}  & \num{-0.0082}  & \num{0.0162}  & \num{0.52}  \\
Percentage of Uninsured Adults          & \num{0.0328}   & \num{0.0075}  & \num{0.0181}   & \num{0.0476}  & \num{<0.01} \\
Median.Household.Income                 & \num{1.3e-04}  & \num{2.7e-06} & \num{1.2e-04}  & \num{1.3e-04} & \num{<0.01} \\
RSI*Region-West                         & \num{-0.00032} & \num{0.01164} & \num{-0.02313} & \num{0.02250} & \num{0.98}  \\
RSI*Region-Northeast                    & \num{-0.013}   & \num{0.014}   & \num{-0.040}   & \num{0.015}   & \num{0.36}  \\
RSI*Region-Midwest                      & \num{-0.0075}  & \num{0.0075}  & \num{-0.0222}  & \num{0.0072}  & \num{0.32}  \\
\bottomrule
\end{tblr}
\end{table}

\paragraph{Model Assessment}\label{model-assessment}
